This is the title [Article v\versionnumber]

### 2022-09-13 manuscript/manuscript.tex

\begin{lstlisting}
parm topology.prmtop
trajin trajectory.nc
gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 \
	gridspacn 0.50 out gist.dat
go
\end{lstlisting}

@@

A minimal PyMOL input to visualize a .dx file is shown here.
For reference, an input structure is loaded from a file named \inlinecode{solute.pdb} (assuming that this file exists).
The oxygen density is visualized at an isolevel of 2 (i.e., twice the reference density).

% The current font makes the O look like a tofu / placeholder
\begin{lstlisting}
load solute.pdb
as surface
load gist-gO.dx, gO
isosurface gO_high, gO, 2.
\end{lstlisting}

@@

Here, we choose not to include any sampling of the conformational states. 
It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly. 
Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
\begin{lstlisting}
parm complex/solvated.parm7
trajin complex/equil/EQUIL-DONE.rst
strip ^2 parmout streptavidin/solvated.parm7
trajout streptavidin/pre-equil.ncrst
\end{lstlisting}
(Adapt for biotin as necessary.)
%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system. \todo{Explain cpptraj rather than our script. The script is for the automated workflow only.}

@@

An example Amber input file for GIST might look like this:
\begin{lstlisting}
restrained 100 ns NpT
&cntrl
  ntx=5, irest=1,
  ioutfm=1,
  ntb=2, iwrap=1,
  ntr=1, restraint_wt=100.0,
    restraintmask="!@H=&!:WAT",
  ntp=1, pres0=1.0, taup=1.0,
  ntc=2, ntf=2,
  ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
  nstlim=50000000, dt=0.002,
  ntwr=50000, ntwx=5000,
  ntpr=5000,
/
\end{lstlisting}
Note that these are very standard MD settings except for the line that defines the restraints.

Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}
pmemd.cuda \
	-O \
	-i 100ns-npt-restraint.in \
	-o md-01.out \
	-p solvated.parm7 \
	-c EQUIL-DONE.rst \
	-x md-01.nc \
	-r md-01.ncrst \
	-ref EQUIL-DONE.rst
\end{lstlisting}

\subsection{Running the GIST analysis for Streptavidin-Biotin Complex}
For easier post-processing, we will use grids that contain the whole molecule.
This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
We also recommend that you center the molecule in space.
For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
\begin{lstlisting}
parm solvated.parm7
trajin md-01.nc 1 1 1
autoimage !(:WAT) origin
strip :WAT
trajout solute-centered.pdb
\end{lstlisting}
Then, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates for each system.
This can be done e.g., using the \software{mdtraj} package in Python.
\begin{lstlisting}[style=python]
import mdtraj as md
pdb = md.load("solute-centered.pdb")
# mdtraj converts to nm.
# *10 to convert back to Angstrom.
min = pdb.xyz[0].min(0) * 10.
max = pdb.xyz[0].max(0) * 10.
print(*list(zip("xyz", min, max)), sep="\n")
\end{lstlisting}
Then, decide on grid dimensions for every system.
For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{15}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+15)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
\todo{ This is handled differently by FindCentroid.py: d = max - min; voxelnr = 
d*5 for x,y,z each; we should handle it similiar or at least mention 
FindCentroid.py here. 
What does gisttools do if the grid is too small? -vah }
We note that this centered grid is usually larger than it needs to be in some directions.
If you want to save some extra memory and calculation time, feel free to align the grid more precisely to the minimum and maximum coordinates, or use a smaller grid that only covers the binding pocket.
You can also rotate the molecule by its principal axes to fit the cuboid grid more exactly.
However, this needs to be done \emph{before} the MD run, since rotating the trajectory damages the periodic box information.

@@

\section{Recent developments}
\subsection{GIST with non-water solvents}
\subsection{GIST with salt-water mixtures}
\section{Cheatsheet}
%This will be a cheatsheet containing all options for cpptrajs GIST implementation
\subsection{Installation}
GIST calculations in cpptraj can be sped up through various levels of 
parallelization. When cpptraj was compiled with the '-cuda' configure flag set, 
the GPU accelerated version of GIST is used automatically.

@@

\textbf{Output}\\
GIST calculations put out a variety of data sorted by voxel into a file 
specified by the "out" keyword. Some of the output data is also automatically 
written to "open data explorer" (.dx) files for convenient visualisation in 
software such as VMD or PyMOL. The following columns can be found in the output 
file:
\begin{tabular}{l l p{.8\textwidth}}
\toprule
Name & Keyword & Description\\
\midrule
gX & X & For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density. If the same element occurs multiple times, the bulk density is scaled accordingly. \\
Esw & X &  Mean solute-water interaction energy density. \\
Eww & X & Mean water-water interaction energy density. \\
PME & [pme] & Water PME energy density. \\
U\_PME & [pme] & Water PME energy density. \\
dTStrans & X & First order translational entropy density. \\
dTSorient & X & First order orientational entropy density . \\
neighbor & X & Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density. \\
dipole & X & Magnitude of mean dipole moment (polarization). \\
order & X & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter. \\
dipolex & X & x-component of the mean water dipole moment density \\
dipoley & X & y-component of the mean water dipole moment density \\
dipolez & X & z-component of the mean water dipole moment density \\
Eij & X & (if \inlinecode{doeij} was specified) Water-water interaction matrix. \\
\bottomrule
\end{tabular}
\end{checklist}
\end{Checklists*}
\section{Checklists}
%Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
%Several useful checklist formats are available, with examples presented in \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates}.
%One example is shown here.

>>>>

\begin{lstlisting}[style=cpptraj]
parm topology.prmtop
trajin trajectory.nc
gist gridcntr 10.5 20.1 30.1 griddim 30 30 30 \
	gridspacn 0.50 out gist.dat
go
\end{lstlisting}

@@

A minimal PyMOL input to visualize a ..dx file is shown here.
For reference, an input structure is loaded from a file named \inlinecode{solute.pdb} (assuming that this file exists).
The oxygen density is visualized at an isolevel of 2 (i.e., twice the reference density).

% The current font makes the O look like a tofu / placeholder
\begin{lstlisting}[style=pymol]
load solute.pdb
as surface
load gist-gO.dx, gO
isosurface gO_high, gO, 2.
\end{lstlisting}

@@

Here, we choose not to include any sampling of the conformational states. 
It is therefore important that the conformations of the monomers in the complex and dissociated states match exactly. 
Use \software{cpptraj} to produce biotin and streptavidin structures based on the complex by stripping the respective other molecule like this:
\begin{lstlisting}[style=cpptraj]
parm complex/solvated.parm7
trajin complex/equil/EQUIL-DONE.rst
strip ^2 parmout streptavidin/solvated.parm7
trajout streptavidin/pre-equil.ncrst
\end{lstlisting}
(Adapt for biotin as necessary.)
%We provide a script called \software{cpptraj\_remove\_mol.sh} that removes a single molecule from a topology and structure using \software{cpptraj} without altering the rest of the system. \todo{Explain cpptraj rather than our script. The script is for the automated workflow only.}

@@

An example Amber input file for GIST might look like this:
\begin{lstlisting}[style=amber-in]
restrained 100 ns NpT
&cntrl
  ntx=5, irest=1,
  ioutfm=1,
  ntb=2, iwrap=1,
  ntr=1, restraint_wt=100.0,
    restraintmask="!@H=&!:WAT",
  ntp=1, pres0=1.0, taup=1.0,
  ntc=2, ntf=2,
  ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
  nstlim=50000000, dt=0.002,
  ntwr=50000, ntwx=5000,
  ntpr=5000,
/
\end{lstlisting}
Note that these are very standard MD settings except for the line that defines the restraints.

Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}[style=bash]
pmemd.cuda \
	-O \
	-i 100ns-npt-restraint.in \
	-o md-01.out \
	-p solvated.parm7 \
	-c EQUIL-DONE.rst \
	-x md-01.nc \
	-r md-01.ncrst \
	-ref EQUIL-DONE.rst
\end{lstlisting}

\subsection{Running the GIST analysis for Streptavidin-Biotin Complex}
For easier post-processing, we will use grids that contain the whole molecule.
This could in theory be avoided for streptavidin and the complex, since only the surrounding of the biotin molecule (i.e., the binding pocket) is required for the analysis.
We also recommend that you center the molecule in space.
For each system (biotin, streptavidin, complex), produce a centered pdb file using \software{cpptraj}:
\begin{lstlisting}[style=cpptraj]
parm solvated.parm7
trajin md-01.nc 1 1 1
autoimage !(:WAT) origin
strip :WAT
trajout solute-centered.pdb
\end{lstlisting}
Then, find the extent (minimum and maximum) of the x-, \mbox{y-,} and z-coordinates for each system.
This can be done e.g., using the \software{mdtraj} package in Python.
\begin{lstlisting}[style=python]
import mdtraj as md
pdb = md.load("solute-centered.pdb")
# mdtraj converts to nm.
# *10 to convert back to Angstrom.
min = pdb.xyz[0].min(0) * 10.
max = pdb.xyz[0].max(0) * 10.
print(*list(zip("xyz", min, max)), sep="\n")
\end{lstlisting}
Then, decide on grid dimensions for every system.
For a system centered at the origin with a spacing of \SI{0.5}{\angstrom} and a buffer of, e.g., \SI{15}{\angstrom} between the solute and the grid wall, you need $4.0\times (d+15)$ bins in the respective direction, where $d$ is the larger of $-$min and $+$max in the respective direction.
\todo{ This is handled differently by FindCentroid.py: d = max - min; voxelnr = 
d*5 for x,y,z each; we should handle it similiar or at least mention 
FindCentroid.py here. 
What does gisttools do if the grid is too small? -vah }
We note that this centered grid is usually larger than it needs to be in some directions.
If you want to save some extra memory and calculation time, feel free to align the grid more precisely to the minimum and maximum coordinates, or use a smaller grid that only covers the binding pocket.
You can also rotate the molecule by its principal axes to fit the cuboid grid more exactly.
However, this needs to be done \emph{before} the MD run, since rotating the trajectory damages the periodic box information.

@@

\section{Recent developments}
\subsection{GIST with non-water solvents}
\subsection{GIST with salt-water mixtures}
In \cite{Waibl2021-gist-salt}, an extension of GIST was presented that can use salt-water mixtures as a solvent.
This allows treating salting-out effects, although it was shown that the salting-out coefficient is over-predicted by GIST unless higher-order entropies are incorporated.
Generally, GIST should be able to treat arbitrary solvent mixtures as long as each molecule is sufficiently rigid.

The \software{cpptraj} implementation of GIST allows the user to specify which components of the system to treat as a solute, using the \inlinecode{solute} keyword.
Everything else will be treated as solvent.
If the solvent contains more than one molecular species, a list of solvent molecule names (i.e., ``\inlinecode{solventmols WAT,NA,CL}'') must be specified.
For each molecule in this list, densities (e.g., \inlinecode{g\_mol\_NA}) and energies (e.g., \inlinecode{Eww\_mol\_NA}) will be computed and written to .dx files.
Note that, e.g., \inlinecode{Esw\_mol\_NA} contains the interactions of every ``\inlinecode{NA}'' molecule with the solute, and \inlinecode{Esw\_mol\_NA} contains the interactions among ``\inlinecode{NA}'' molecules as well as their interactions with all other solvents, divided by 2 to account for double counting.
Furthermore, the first molecule in this list will be treated as the main solvent, for which first-order entropy values will be computed.

For instance, a GIST calculation in salt water, including the first-order water entropy, can be run as follows:

\begin{lstlisting}[style=cpptraj]
gist gridcntr <x> <y> <z> \
griddim <Nx> <Ny> <Nz> gridspacn <val> \
solute !(:WAT,NA,CL) solventmols WAT,NA,CL
\end{lstlisting}

The current implementation of the entropy calculation in \software{cpptraj} requires at least 3 atoms in the main solvent.
To compute the first-order entropy of the ions, as well as an approximate second order entropy, Python code is available on \url{https://github.com/liedllab/second-disorder}.

\section{Cheatsheet}
%This will be a cheatsheet containing all options for cpptrajs GIST implementation
\subsection{Installation}
GIST calculations in cpptraj can be sped up through various levels of 
parallelization. When cpptraj was compiled with the '-cuda' configure flag set, 
the GPU accelerated version of GIST is used automatically.

@@

\textbf{Output}\\
GIST calculations put out a variety of data sorted by voxel into a file 
specified by the "out" keyword. Some of the output data is also automatically 
written to "open data explorer" (.dx) files for convenient visualisation in 
software such as VMD or PyMOL. The following columns can be found in the output 
file:
\begin{tabular}{l l p{.8\textwidth}}
\toprule
Name & Keyword & Description\\
\midrule
gX & X & For every element in the main solvent, the number density of atoms found in the voxel, in units of the bulk density. If the same element occurs multiple times, the bulk density is scaled accordingly. \\
Esw & X &  Mean solute-water interaction energy density. \\
Eww & X & Mean water-water interaction energy density. \\
PME & [pme] & Water PME energy density. \\
U\_PME & [pme] & Water PME energy density. \\
dTStrans & X & First order translational entropy density. \\
dTSorient & X & First order orientational entropy density . \\
neighbor & X & Mean number of waters neighboring the water molecules found in this voxel multiplied by the voxel number density. \\
dipole & X & Magnitude of mean dipole moment (polarization). \\
order & X & (if \inlinecode{doorder} was specified) Average Tetrahedral Order Parameter. \\
dipolex & X & x-component of the mean water dipole moment density \\
dipoley & X & y-component of the mean water dipole moment density \\
dipolez & X & z-component of the mean water dipole moment density \\
Eij & X & (if \inlinecode{doeij} was specified) Water-water interaction matrix. \\
\bottomrule
\end{tabular}
\end{checklist}
\end{Checklists*}

\section{Checklists}
%Tutorials do not necessarily require the use of a checklist as in Best Practices documents; however, they can include these if desired.
%Several useful checklist formats are available, with examples presented in \texttt{sample-document.tex} in \url{github.com/livecomsjournal/article_templates/templates}.
%One example is shown here.
